---
title: "Clustering UFC Fighters: K-means and Hierarchical"
author: "Jason Willett"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@ Read in all the data sets
```{r}
ufc = read.csv("ufcdataupdated.csv") # complete dataset of all fights
ufcfighters = read.csv("ufcfighters.csv") # physical characteristics of all fighters
bigufc = read.csv("bigufc.csv") # unpolished dataset of fights
```

## Of the 2697 fights:
## Red won 56% of the fights.
## Blue won 42% of the fights.
## Draw occurred in 2% of the fights.

## Create a binary win column
```{r}
Binwin = c(rep(0, length(ufc$Winner)))
for (i in 1:length(ufc$Winner)) {
  if (ufc$Winner[i] == "Red"){
    Binwin[i] = 1
  }
  
}
ufc = data.frame(cbind(ufc,Binwin))
```
## Now I have added an entire column that indicates whether or not Red fighter won. It is binary. ## 1 means won and 0 means loss/draw.


# Create a dataframe of all the fighters featured within the fights
```{r}
rlevs = c(levels(factor(ufc$R_fighter))) # find unique red fighters
blevs = c(levels(factor(ufc$B_fighter))) # find unique blue fighters
combinedlevs = c(rlevs,blevs)
levsframe = data.frame(c(rep(0,length(combinedlevs))))
for (i in 1:length(combinedlevs)) {
  levsframe[i,1] = combinedlevs[i]
  
}
```
## Factor all the rows of levsframe to find more levels
```{r}
levsframe[,1] = factor(levsframe[,1])
```
## Find unique fighters for combined rlevs and blevs frame
```{r}
finnamelevs = data.frame(c(levels(levsframe[,1])))
```
## Change type of element from factor to character
```{r}
finnamelevs[,1] = as.character(finnamelevs[,1])
```

## Load in tidyr package
```{r}
library(tidyr)
```
## Remove unnecessary columns from ufc dataset: rowidx, referee name, location, Winner (created ## binary column)
```{r}
ufc = subset(ufc, select = -c(X, Unnamed..0,Referee,location,Winner))
```
## Load dplyr package
# Keep only columns relating to red fighters
```{r}
library(dplyr)
R_col = ufc %>% select(contains("date"),,contains("weight_class"),contains("R_"))
```
## Keep only columns relating to blue fighters
```{r}
B_col = ufc %>% select(contains("date"),contains("weight_class"),contains("B_"))
```
## Remove "R_" and "B_" from column names to standardize 
```{r}
names(R_col) = gsub("R_","",names(R_col))
names(B_col) = gsub("B_","",names(B_col))
```
## Remove any remaining columns that should not be in the dataset
```{r}
R_col = subset(R_col, select = -c(B_avg_SIG_STatt,B_avg_SIG_STlanded,B_avg_SIG_STpct,B_avg_TOTAL_STatt,B_avg_TOTAL_STlanded,B_avg_opp_SIG_STatt,B_avg_opp_SIG_STlanded,B_avg_opp_SIG_STpct,B_avg_opp_TOTAL_STatt,B_avg_opp_TOTAL_STlanded, B_win_by_TKO_Doctor_Stoppage,age))
B_col = subset(B_col, select = -c(R_avg_SUATT,R_avg_opp_SUATT, age))
```
## Remove any antecedents to standardize
```{r}
names(R_col) = gsub("B_","",names(R_col))
names(B_col) = gsub("R_","",names(B_col))
R_col[1,]
```
## Combine red and blue revised subsetted data into one dataset
```{r}
ufc = rbind(R_col,B_col)
```
## Find most recent fight for each fighter
```{r}
ufc = ufc[order(ufc$date),] #prder the fights by date
recfight = ufc[which(ufc$fighter == finnamelevs[1,]),]
```
## Find most recent for each fighter and append to recfight dataframe
```{r}
for (i in 2:length(finnamelevs[,1])) {
  maxx = max(which(ufc$fighter == finnamelevs[i,]))
  fight = ufc[maxx,]
  recfight = rbind(recfight,fight)
  
}
recfight = data.frame(recfight) # make dataframe
```
## Make rownames the fighters name for making it clear in clustering later on
```{r}
rownames(recfight) = recfight$fighter
```
## Remove any remaining columns that will not be necessary for clustering or clumns that are not ## of interest
```{r}
fighter_features = subset(recfight, select = -c(losses,odds, Stance, wins,total_rounds_fought,total_time_fought.seconds.,total_title_bouts,win_by_Decision_Majority,win_by_Decision_Split,win_by_Decision_Unanimous,win_by_KO.TKO,win_by_Submission,win_by_TKO_Doctor_Stoppage,date,fighter,current_lose_streak,current_win_streak,draw,weight_class,longest_win_streak))
```

## Scale and standaridize all columns to make sure no feature is given more power because its on a larger scale
```{r}
m = apply(fighter_features,2,mean)
s = apply(fighter_features,2,sd)
fighter_features = scale(fighter_features,m,s) # z-score standardization
```
## Find sum of WCC for each number of clusters allowed to perform elbow method
```{r}
set.seed(1234)
kmean2 = kmeans(fighter_features,2)
mWCC_list = mean(kmean2$withinss)
for (i in 2:10) {
  kmea = kmeans(fighter_features, i)
  mWCC = mean(kmea$withinss)
  mWCC_list = c(mWCC_list,mWCC)
}
mWCC_list
```


## Create plot of sWCC to perform elbow method
```{r}
cluster_size = c(2:10)
plot(cluster_size,mWCC_list[2:10], main = "Elbow Method for Determining Number of Clusters", xlab = "Number of Clusters", ylab = "Mean of WC Sum of Squares", type = "o")
```
## Based off elbow method, looks like 3 or 4 clusters is the optimal number of clusters

## Find silhouette coefficent: alternate method for finding optimal number of clusters
```{r}
library(cluster)
set.seed(2020)
asw <- numeric(20)
for (k in 2:20){
  asw[k] <- pam(fighter_features, k) $ silinfo $ avg.width
  k.best <- which.max(asw)
}
cat("silhouette-optimal number of clusters:", k.best, "\n")
```


## According to the elbow method, it appears we should set k=3. However, based off 
## SilhouetteCoefficients, we should set k=2.

## Use k-means with 2 clusters and 3 clusters 
```{r}
set.seed(2020)
results2 = kmeans(fighter_features, 2)
results3 = kmeans(fighter_features, 3)
```
## Explore the k-means with 3 clusters
```{r}
table(recfight$weight_class,results3$cluster) # table comparing cluster assignment to weight class
plot(recfight$Weight_lbs,recfight$age,col=results3$cluster) # age vs cluster
plot(recfight$Height_cms,recfight$Reach_cms,col=results3$cluster) # arm length
# plots below are some fighting statistic comparisons
plot(recfight$avg_SIG_STatt,recfight$avg_SIG_STlanded,col=results3$cluster) 
plot(recfight$avg_HEAD_att,recfight$avg_HEAD_landed,col=results3$cluster)
plot(recfight$avg_DISTANCE_landed,recfight$avg_opp_DISTANCE_att,col=results3$cluster)
plot(recfight$avg_opp_GROUND_att,recfight$avg_opp_GROUND_landed,col=results3$cluster)
```
## Create standardized plots for fighter stats and opponent stats
```{r}
mha = mean(recfight$avg_HEAD_att)
sha = sd(recfight$avg_HEAD_att)
avg_head_att = scale(recfight$avg_HEAD_att,mha,sha)
mhl = mean(recfight$avg_HEAD_landed)
shl = sd(recfight$avg_HEAD_landed)
avg_head_land = scale(recfight$avg_HEAD_landed,mhl,shl)
mssa = mean(recfight$avg_SIG_STatt)
sssa = sd(recfight$avg_SIG_STatt)
avg_SS_att = scale(recfight$avg_SIG_STatt,mssa,sssa)
moha = mean(recfight$avg_opp_HEAD_att)
soha = sd(recfight$avg_opp_HEAD_att)
avg_opp_head = scale(recfight$avg_opp_HEAD_att,moha,soha)
moda = mean(recfight$avg_opp_DISTANCE_att)
soda = sd(recfight$avg_opp_DISTANCE_att)
avg_opp_dist = scale(recfight$avg_opp_DISTANCE_att,moda,soda)
plot(avg_head_att,avg_SS_att,main = "Fighter Standardized Averages", xlab = "Head Attempts", ylab = "Significant Stike Attempts", col = results3$cluster)
plot(avg_opp_dist,avg_opp_head, main = "Opponent Standarized Averages",xlab = "Opponent Distance Attempts",ylab = "Opponent Head Attempts",col = results3$cluster)

```




## Find win probability for each cluster
```{r}
cluster1wper = sum(recfight$wins[which(results3$cluster == 1)]) / (sum(recfight$wins[which(results3$cluster == 1)]) + (sum(recfight$losses[which(results3$cluster == 1)])) + (sum(recfight$draw[which(results3$cluster == 1)])))
cluster2wper = sum(recfight$wins[which(results3$cluster == 2)]) / (sum(recfight$wins[which(results3$cluster == 2)]) + (sum(recfight$losses[which(results3$cluster == 2)])) + (sum(recfight$draw[which(results3$cluster == 2)])))
cluster3wper = sum(recfight$wins[which(results3$cluster == 3)]) / (sum(recfight$wins[which(results3$cluster == 3)]) + (sum(recfight$losses[which(results3$cluster == 3)])) + (sum(recfight$draw[which(results3$cluster == 3)])))
winp = c(cluster1wper,cluster2wper,cluster3wper)
winp
```
## Cluster 1 has a significantly higher win proportion than Cluster 2. Cluster 3 has a 
## significantly higher win proportion than Cluster 2. Cluster 1 and Cluster 3 are not 
## significantly different. 
## Significane was found from T-test

## Hierarchal Clustering
```{r}
set.seed(2020)
distances = dist(fighter_features)
hclustermod2 = hclust(distances, method = "average")
plot(hclustermod2, hang = -4, cex=.01)
hclustermod = hclust(distances)
plot(hclustermod, hang = -4, cex=.01)
```

```{r}
set.seed(2020)
clusterCut1 = cutree(hclustermod, 3)
table(recfight$weight_class,clusterCut1)
```
```{r}
#set.seed(2020)
#clusterCut2 = cutree(hclustermod2, 3)
#table(recfight$weight_class,clusterCut2)
```
Average Hierarchal did not do well.

## Understanding clusters from k-meabs with 3 clusters (Optimal Clusters)
```{r}
results3
```



